Interpretable Neural Networks using EAGGA

### 2025-03-15 paper/main.tex
finished own extension

\subsection{Overall Algorithm}
We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
\end{itemize}
For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only these features can interact with each other.

@@

than the multiplicative one happening for instance in a linear model.

@@

where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to

@@

$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l-1)}\times p^{(l)}}$ where $\mathbb{R}_{0+}=[0,\infty)$.
$\boldsymbol{o}_{\text{out}}^{(l)}=ReLU(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ Is therefore a monotonically increasing

@@

\subsection{Experimental Setup}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{./dataset_split.png}
  \caption{TODO: caption}
  \label{fig-dataset-split}
\end{figure}
  * evaluation, holdout, cv, early stopping
    - evaluation via dominated hypervolume along AUC-ROC, NF, NI, NNM as defined in original paper
      * NF simply rel. \# of included features
      * NI = sum of all possible pairwise interactions in each group over all possible pairwise interactions among all features = $\frac{\sum_g^G {p_g \choose 2}}{{p \choose 2}}$
      * NNM = rel. \# unconstrained features
    - outer holdout split: 2/3 train, 1/3 test (as in paper)
      * run EAGGA on holdout train split, i.e. train + select best $\mu=100$ individuals based on non-dom-sorting (ascending) + crowding distance (descending) as tie breaker
    - inner CV split on outer train portion: 5-fold (as in paper)
      * in each fold fit model on 80\% of CV-train portion, use remaining 20\% of CV-train for early stopping
    - early stopping criterion
      * for each fold, always train for min 200 epochs, keep track of model with lowest loss
      * after that, use patience of 100: if current model's loss is > mean of last 100 epochs' losses -> stop early
      * if no early stopping, train each fold for max 10 minutes
      * then stop training, return model with lowest loss
    -> refere Figure \ref{fig-dataset-split}
    - then evaluate on last remaining from CV + keep best $\mu$ individuals + generate $\lambda=10$ offspring for new generation + repeat
    - final evaluation of individuals of pareto set
      * for each individual, train model on training set of hold-out for max \# of epochs taken during CV-training
      * decided on max \# epochs instead of mean after looking at loss graphs in preliminary experiments, also refer Figure \ref{fig-es-losses}
      * those show that there is no uptick in losses on the early stopping portion (disjunct from training portion) of the set, hence max is reasonable
  * hardware: training on Sagemaker Notebook instance
    - initial experiments on ml.g4dn.xlarge instance (2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/)
      using cuda not much faster than on ml.t3.medium (2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/)
    - thus decided for more economic + ressourcen-schonend t3.medium
  * did not train on philippine and gina datasets (308, 970 features, respectively), as they ran out of memory when computing interaction detectors
  * known bugs
    - in rare cases (anecdotally once every ~5-10 datasets), gga\_mutate seems to be generating -1 as monotonicity attribute, despite np.random.randint(low=0, high=2, size=1)
      * loop crashes at group\_structure creation, for these cases subtract previous runtime from 8hrs + load from last generation in output via load\_population
      * so far only happened for madeline in after gen-8.json after 5hrs 45mins (at start of gen-9, which wasn't exported yet) -> loaded this and ran for another 2hrs 15mins
  * only 18 datasets

>>>>

\subsection{Overall Algorithm}
We implement EAGGA largely as described by \citet{EAGGA} with only minor changes due to no straightforward transferability either due to different implementations
of some classes and functions in R compared to Python or differences in model properties between XGBoost and neural nets.\\
% TODO: after architectural details:
Our NNs use the following hyperparameters initialised as described
\begin{itemize}
  \item Total number of layers $\in\{3, ..., 10\}$, initialised from a truncated Geometric distribution with $\pi=0.5$.
        This includes input and output layers, the number of hidden layers is thus $\in\{1, ..., 8\}$.
  \item Number of nodes per hidden layer $\in\{3, ..., 20\}$, initialisation from a truncated Geometric distribution with $\pi=0.5$.
  \item Dropout probability $\in[0, 1]$, initialised from a truncated Gamma distribution with shape 2 and scale 0.15.
\end{itemize}
For the group structure initialization we also employ a feature, interaction, and monotonicity detector.
\\
The feature detector is as described by \citet[p. 542]{EAGGA}, but instead of fitting ten trees and using the relative number of features used as
probability for the truncated Geometric distribution, we use a constant value of 0.5, as the sklearn DecisionTree examination is not straightforward.
Further, in preliminary experiments we found the sampled number of features to occassionally be larger than the number of non-zero values in the
normalised information gain filter. This can happen if the former is equal to the total number of features and one or more features are independent form the
target. In these edge cases we use all features with non-zero filter-values instead.
\\
The interaction detector similarly uses a constant probability of 0.5 to sample the number of interactions to be used instead of the relative number of interactions
from ten decision trees, for the same reasoning as for the feature detector. Also, for the FAST algorithm computing the interaction scores we don't use the residual
sum of squares but the mean accuracy, as the datasets all have binary targets and FAST thus fits a logistic instead of a linear regression.
\\
The monotonicty detector is implemented as described by \citet[p. 543]{EAGGA}, using the default decision tree hyperparameters from mlr3 in Python.
\\
Both the models in the interaction and the monotonicity detector (FAST and an ensemble of decision trees, respectively) use 80\% of the data for model estimation and
the remaining 20\% to compute their scores.

@@

Each equivalence class in the group structure has its own sub-network, so that it is guaranteed that only the features within that group can interact with each other.

@@

than the multiplicative one happening, for instance, in a linear model.

@@

where $\boldsymbol{o}_{\text{out}}^{(l)}$ denotes the value of layer $l$ after applying ReLU, $\boldsymbol{W}\in\mathbb{R}^{p^{(l)} \times p^{(l-1)}}$ refers to

@@

$\boldsymbol{W}\in\mathbb{R}_{0+}^{p^{(l)}\times p^{(l-1)}}$, where $\mathbb{R}_{0+}=[0,\infty)$.
Therefore, $\boldsymbol{o}_{\text{out}}^{(l)}=\text{ReLU}(\boldsymbol{o}_{\text{in}}^{(l)})=\max(0,\boldsymbol{o}_{\text{in}}^{(l)})$ is a monotonically increasing

@@

\subsection{Experimental Setup}
  \includegraphics[width=\linewidth]{./dataset_split.png}
  \label{fig-dataset-split}
We evaluate our models just like \citet[pp. 543f]{EAGGA} by computing the dominated hypervolume along AUC, NF, NI, NNM with reference point $(0, 1, 1, 1)^T$.
Internally, this is implemented as minimising $1-\text{AUC}$ and a reference point $(1, 1, 1, 1)^T$.
For NF, we simply count the relative number of included features.
NI sums the numbers of all possible pairwise interactions in each group $\sum_g^G {p_g \choose 2}$ and divides them by the number of all possible pairwise interactions
on the entire dataset ${p \choose 2}$.
NNM counts the relative number of features in unconstrained equivalence classes.
Our dataset is split as visualised in Figure \ref{fig-dataset-split}.
We apply an outer holdout train-test split with 1/3 of the dataset being held for final testing, the 2/3 are used for 5-fold cross-validated training of each individual,
as described by \citet[p. 543]{EAGGA}.
What is different for our implementation is that on each training portion of the CV split, we further apply a holdout split, where 80\% are used for model training and
20\% are reserved for early stopping.
We use a conservative early stopping stratetgy, where each fold is always trained for a minimum of 200 epochs, keeping track of the fold's model with the lowest loss.
After 200 epochs, our criterion with a patience of 100 epochs is applied: now, if the current epoch's model loss is greater than the mean of the last 100 epochs'
losses, training in this fold stops early and returns the model with the lowest loss over the entire fold's training period.
If no early stopping is triggered, we train each fold for a maximum of 10 minutes and return the model with the lowest loss on the early stopping portion.
The model is then evaluated on the remaining portion from the CV split and after each generation, we keep the best $\mu=100$ individuals and generate $\lambda=10$
offspring for the next generation and repeat this for 8 hours.
The final evaluation of the individuals from the cross-validated Pareto set is then conducted on the held 1/3 of the dataset. Each individual
is trained on the 2/3 training portion for the maximum number of epochs the individual required in the cross-validated training.
Oftentimes, for cross-validation with early stopping, the final model is only trained on the mean or median number of epochs from CV.
We decided specifically on the maximum after examining the loss over epochs in preliminary experiments, as plotted in Figure \ref{fig-es-losses}
for the final experimental setup. Those show that there is no uptick in losses on the early stopping portion indicating that training too long does not increase
the validation error.
We trained in Sagemaker Notebooks on AWS and did initial training runs on
ml.g4dn.xlarge\footnote{2 vCPUs, 16GiB RAM, 1 NVIDIA T4, cf. https://aws.amazon.com/de/ec2/instance-types/g4/} instances with CUDA.
There was, however, no noticeable speed-up in model training compared to the much more economical
ml.t3.medium\footnote{2 Intel Xeon 8000 vCPUs, 4GiB RAM, cf. https://aws.amazon.com/de/ec2/instance-types/t3/} instances, after which we decided to run the
final experimental setup on those.
In total, we run our models on the same datasets as \citet[p. 544]{EAGGA}, except for the 2 largest ones ``philippine'' and ``gina'' (308 and 970 features, respectively),
as those ran out of memory when initializing the interaction effects.
In rare cases (anecdotally once every ~10 datasets), we observed our monotonicity detector generating a group monotonicty attribute of -1, which crashed the computation.
For these cases, we implemented a function to import a previous generation of individuals to be able to pick up from where the training crashed and not train anew.
This happened once for the dataset ``madeline'' after generation 9 and training for 5 hours 45 minutes, and hence we imported the generation and ran it for the remaining
2 hours 15 minutes.
